---
title: "5123 Final Prpject"
name: "Jianjian Wu"
output: html_document
date: "2026-04-09"
---

############################## EPSY5123 Jianjian's Project ##############################

### Research Question
The main research question of this project is: 
To what extent are family-related factors associated with students’ Portuguese language performance?

### Data Source
The dataset I used is the public Portuguese language dataset in UCI Student Performance repository, which included the student records from two Portuguese secondary schools. The dataset contains students’ Portuguese language grades and background information. The full dataset includes 649 cases, 30 predictor features regarding demographic, family, social, and school-related variables and Portuguese language score records. 

#*original data source: https://archive.ics.uci.edu/dataset/320/student+performance 
(click the link and you can find the dataset introduction information)




### Database preparation

##Loading the packages and checking the original data.

```{r load-packeage-and-data}

# loading packages
library(ggplot2)
library(tidyverse)
library(dplyr)

# loading and checking data 
student.por_cleaned <- read.csv("student-por_cleaned.csv")
head(student.por_cleaned)
glimpse(student.por_cleaned)

```

##Revising the variable type of the dataset

After checking the original dataset, I find that all the variables are in "character" type, and this is impossible for me to do some statistical analysis with these data. 

So I need to correct the variable types.

After data revision, we know that in this dataset, there are 649 rows and 33 columns, which means we have 649 cases and 33 variables. 

```{r dataset-revision}

# data backup
student <- student.por_cleaned 
student 
glimpse(student) 

# changing variables from character into numeric ones
student$age <- as.numeric(student$age)
student$Medu <- as.numeric(student$Medu)
student$Fedu <- as.numeric(student$Fedu)
student$traveltime <- as.numeric(student$traveltime)
student$studytime <- as.numeric(student$studytime)
student$failures <- as.numeric(student$failures)
student$famrel <- as.numeric(student$famrel)
student$freetime <- as.numeric(student$freetime)
student$goout <- as.numeric(student$goout)
student$Dalc <- as.numeric(student$Dalc)
student$Walc <- as.numeric(student$Walc)
student$health <- as.numeric(student$health)
student$absences <- as.numeric(student$absences)
student$G1 <- as.numeric(student$G1)
student$G2 <- as.numeric(student$G2)
student$G3 <- as.numeric(student$G3)

# changing variables from character into factors
student$sex <- as.factor(student$sex)
student$address <- as.factor(student$address)
student$famsize <- as.factor(student$famsize)
student$Pstatus <- as.factor(student$Pstatus)
student$Mjob <- as.factor(student$Mjob)
student$Fjob <- as.factor(student$Fjob)
student$guardian <- as.factor(student$guardian)
student$schoolsup <- as.factor(student$schoolsup)
student$famsup <- as.factor(student$famsup)
student$paid <- as.factor(student$paid)
student$activities <- as.factor(student$activities)
student$nursery <- as.factor(student$nursery)
student$higher <- as.factor(student$higher)
student$internet <- as.factor(student$internet)
student$romantic <- as.factor(student$romantic)

# check the dataset: 649 rows and 33 columns
glimpse(student) # now we have done the data preparation for analysis

```


### Variable identification and basic data wrangling

##Selecting variables

According to the database introduction, we know that G3 is "final grade", and it is the outcome variable of the current study. Next, I need to identify some potential family-related predictors of the outcome variable. 

I chose the following 5 family-related factors as predictors:

*"famsize" (family size, 'LE3' - less or equal to 3; 'GT3' - greater than 3)
*"Medu" (mother's education, 0 - none; 1 - primary education (4th grade); 2 - 5th to 9th grade; 3 - secondary education; 4 - higher education)
*"Fedu" (father's education, 0 - none; 1 - primary education (4th grade); 2 - 5th to 9th grade; 3 - secondary education; 4 - higher education)
*"famsup" (family educational support, yes or no))
*"famrel" (quality of family relationships, from 1 - very bad to 5 - excellent)

```{r predictor-identification}

# select the 6 related variables and save them into a smaller dataset for this study 
student_fam <- student %>%
  select(G3, famsize, Medu, Fedu, famsup, famrel) 

glimpse(student_fam) # check the database I will use latter ("student_fam")

```

##Checking mising values with filter() function

After checking, we can see that there is no missing value in this dataset.

```{r NA-check}
# get rid of the missing values
student_fam <- student_fam %>%
  filter(
    !is.na(G3),
    !is.na(famsize),
    !is.na(Medu),
    !is.na(Fedu),
    !is.na(famsup),
    !is.na(famrel))

nrow(student_fam) # still 649 rows
```


### Narrative description and Visualization

I explored the following sub-questions below:

# Q1: Generally describe the mean score and distribution of G3 among 649 students.
- The average G3 score is 11.91(SD = 3.23). 
- If we categorize the G3 score as three levels, that is low(0-8),middle(8-14) and high(15-20), we can see that there were 30 students in low_level group, 488 students in middle_level group, and 131 students in high_level group.Therefore, more than half of the students got middle_level G3 scores.

```{r basic-desctiption}

# Descriptive statistics of G3 
summary(student_fam$G3)
mean(student_fam$G3, na.rm = TRUE) # mean G3 score: 11.91
sd(student_fam$G3, na.rm = TRUE) # sd of G3 score: 3.23

# create a new variable "grade_level" based on G3: low, middle and high grade
student_fam <- student_fam %>% 
  mutate(
    grade_level = case_when(
      G3 <8 ~ "low", # G3 score is lower than 8, low level
      G3 <15 ~ "middle", # G3 score is between 8-14, middle level
      G3 >=15 ~ "high" ) # G3 score is 15 or higher, high level
  )

# the distribution of grade levels
glimpse(student_fam)  # check the database
table(student_fam$grade_level) # high_level:131; low_level:30; middle_level:488

```


# Q2: Visualize the general distribution of students' G3 score.

```{r exploratory-plot}

#a histogram plot of overall G3 distribution
ggplot(student_fam, aes(x= G3))+
  geom_histogram(binwidth = 1, fill = "pink", color = "white")+
  labs(
    title = "Distribution of Students' Final Grade",
    x = "G3 Score",
    y = "Numbe of Students"
  )

#a bar chart plot to display the distribution of G3 in different levels
ggplot(student_fam, aes(x= grade_level))+
  geom_bar(fill = "pink")+
  labs(
    title = "Distribution of G3 by Grade Level",
    x = "G3 Score by Grade Level",
    y = "Numbe of Students"
  )

```


# Q3: Are there any difference in G3 between students with family support and students without family support?
- Yes. The group of students with family support got a little higher average G3 score than the group of students without family support did.
- After grouping, we can see the mean G3 score of no-family-support group(n=251) is 11.67(sd=3.60), while that of have-family-support group(n=398) is 12.96(sd=2.97). 

```{r famsup-comparison}

# compare the G3 between students with family support or and students without family support
student_famsup <- student_fam %>% 
  group_by(famsup) %>% # group the students by family support variable
  summarize(
  mean_G3 = mean(G3, na.rm = TRUE),
  sd_G3 = sd(G3, na.rm = TRUE),
  n = n())

student_famsup 
# famsup_no group(n=251), mean_G3=11.67(sd=3.60)
# famsup_yes group(n=398), mean_G3=12.07(sd=2.97)
```


# Q4: Using a plot to investigate whether students from larger family(GT3) differ in G3 scores compared to those from smaller families(LE3).
- According to the plot, we can see that the median scores for GT3 group(family size more than 3) and LE3 group(family size less than3) are very similar, this suggests that family size does not be strongly associated with students' Portuguese reading performance. However, there is a litter more low-score outliers in GT3 group, this may indicate that students from larger families are likely to get lower scores in Portuguese reading performance.

```{r family-size-and-G3}
ggplot(student_fam, aes(x = famsize, y = G3))+
  geom_boxplot(fill = "lightblue")+
  labs(
    title = "G3 by Fanily Size",
    x = "Family Size (GT3 vs. LE3)",
    y = "G3 Score"
  )
```

# Q5: How is the distribution of parents' education levels among the 649 students?
- The distribution of parents’ education levels differs slightly between mothers and fathers.
- Mother's education levels are more evenly distributed across different levels, with a relatively large proportion at the highest level(level 4, n=175) and the largest group at level 2 (n=186). This suggests that a considerable number of mothers have relatively high educational attainment.
-Fathers’ education levels concentrated more in the middle levels, particularly level 2(n= 209) and level 1(n=174), with fewer fathers at the highest level(level 4, n=28).

```{r parent-edu-plot}
#count the number of students by mother's and father's education level
table(student_fam$Medu) # "0"-6; "1"-143; "2"-186,"3"-139; "4"-175
table(student_fam$Fedu) # "0"-7; "1"-174; "2"-209,"3"-131; "4"-128

# a plot for mother education level
ggplot(student_fam, aes(x = Medu))+
  geom_bar(fill = "pink")+
  labs(
    title = "Mother Education Level Distribution"
  )

# a plot for father education level
ggplot(student_fam, aes(x = Fedu))+
  geom_bar(fill = "lightblue")+
  labs(
    title = "Father Education Level Distribution"
  )
```


# Q6: Are there any difference in G3 for students with different parent education levels?
- Yes. From the plots, we can see some differences in G3 scores across different parent education levels.
- In general, students with higher parent education are likely to get higher G3 scores. Yet the differences are small.
- We can also see that the many students whose parents just had low education level, but the students can still get middle level of average G3 score. This suggests that parents education level may not totally predict student reading performance.

```{r parent-edu-compare}
# show the mean G3 grouped by mother education of different levels
medu_mean <- student_fam %>%
  group_by(Medu) %>%
  summarize(
    mean_G3 = mean(G3),
    n = n()
  )

# show the mean G3 grouped by father education of different levels
fedu_mean <- student_fam %>%
  group_by(Fedu) %>%
  summarize(
    mean_G3 = mean(G3),
    n = n()
  )

# visualize mother edu level and G3
ggplot(medu_mean, aes(x = Medu, y = mean_G3))+
  geom_bar(stat = "identity",fill = "pink")+ # show G3 mean score, not count student number
  labs(
    title = "Average G3 Score by Mother Edu Level",
    x = "Mother Education Level",
    y = "Average G3 Score"
  )

# visualize father edu level and G3
ggplot(fedu_mean, aes(x = Fedu, y = mean_G3))+
  geom_bar(stat = "identity",fill = "lightblue")+ # show G3 mean score, not count student number
  labs(
    title = "Average G3 Score by Father Edu Level",
    x = "Father Education Level",
    y = "Average G3 Score"
  )
```

### Modeling

On the basis of the modeling results, we know that this model is statistically significant (F = 9.63, p<.001), but the explanatory power is relatively low (R-squared=0.07), indicating that the model can explain approximately 7% of the variance.

Among the predictors, only mother’s education level (Medu) shows a significant positive association with G3 (p<.001). The relationship among other family-related predictors and G3 are not significant. 

This suggests that among all the selected family-related predictors, mother's education level is the only one which is positively related to adolescents' Portuguese reading performance.

```{r modeling}
# modeling with 5 predictors
model1 <- lm(G3~famsize + Medu + Fedu + famsup + famrel,
             data = student_fam) # intercept= 9.0397

summary(model1)
```

### Creating Function for Famrel

In this section, I would like to create a function for categorizing the quality of family relationships. When the score of "famrel" <3, it will be labeled as "bad". When 3 < "famrel" < 5, it will be labeled as "good". 

```{r adding-function}
famrel_level <- function(famrel) {
  if(famrel >= 3) {
    return("high")
  }else {
    return("low")
  }
}

famrel_level(4) # "high"
famrel_level(2) # "low"
```



